Smart Model Elimination for Efficient Automated Machine Learning

### 2024-06-21 paper/paper.tex
Update paper.tex

\begin{table}
  \caption{Survey of Existing AutoML Frameworks}
  \label{automl-survey}
  \centering
  \begin{tabular}{lllllllll}
    \toprule
    Framework & EOU & Algs & FE & TM & HPT & Inter & Custom\\
    \midrule
    H2O AutoML & Easy & 10 & Basic & Yes & Auto & SHAP & High\\
    TPOT & Hard & 10 & Basic & No & Genetic & POJO & Moderate\\
    MLJAR & Easy & 10 & Advanced & Yes & Auto & Basic & High\\
    FLAML & Easy & 10 & Limited & Yes & Auto & SHAP & Low-Moderate\\
    LightAutoML & Easy & 10 & Advanced & Yes & Auto & Basic & Moderate\\
    GAMA & Hard & 10 & Basic & Yes & Auto & SHAP & Moderate\\
    \bottomrule
  \end{tabular}
  \captionsetup{font=footnotesize}
  \caption*{\textbf{Abbreviations}: EOU: Ease of Use, Algs: Algorithms, FE: Feature Engineering, TM: Time Management, HPT: Hyper-Parameter Tuning, Inter: Interface, Custom: Customization}
\end{table}
Table 1: Comparison of six AutoML frameworks based on commonmly evaluated criteria as well as more obscure features relevant to the medical diagnosis process

@@

To implement this model elimination process, we trained a boosting model on 268 kaggle datasets. We choose datasets that were medically related and focused our approach to binary classification. We extracted the attributes in Table 2 from each dataset, this is the input for the boosting model. We then ran a "dumb" framework that trained every model on each dataset. The accuracy of every model is ranked and every model is treated as a label with its value being its rank. Then we trained a multilabel regression to predict the rank of each model giving the extracted attributes. We produced a median spearman correlation of 0.7301 between the predicted rank and the actual rank and a median p-value of 1.032e-05. Furthermore, we guarantee that we can predict a top two model within the predicted top ten 90.12\% of the time.

When a user inputs a dataset into SMEML, it first extracts the attributes from the dataset. It feeds these attributes into the boosting model to get the predicted rank of each model. Since we have determined that we can predict a top two model within the predicted top ten quite confidently we can eliminate the bottom 18 models from the list. We run bayseian optimization on the top ten models to get the best hyperparamters for each model. We select the top models that have significantly better performance than other models. These models are stacked together to form the final model. This process is shown in Figure 1.
\begin{table}
  \caption{Attributes extracted from a dataset}
  \label{model-options-table}
  \centering
  \begin{tabular}{lll}
    \toprule
    Attribute & Formula & Purpose\\
    \midrule
    \# of Rows & $|\mathbf{A}|_r$ & Data Size \\
    \hline
    \# of Columns & $|\mathbf{A}|_c$ & Data Size \\
    \hline
    Target Distribution &
    $\frac{\max(\text{count}_1, \text{count}_0)}{|\mathbf{A}|_r}$
    & Data Type \\
    \hline
    Numerical Columns &
    $\frac{|\mathbf{A}_{num}|}{|\mathbf{A}|_{c}}$
    & Data Type \\
    \hline
    Binary Categorical Columns &
    $\frac{|\mathbf{A}_{bin}|}{|\mathbf{A}|_{c}}$
    & Data Type \\
    \hline
    Mean \# of Distinct Values  & $\mu \left\{\text{\# of distinct values in column}\right\}$ & Data Type \\ 
    \hline
    Mean IQR  & $\mu \left\{\text{IQR of column}\right\}$ & Data Distribution \\ 
    \hline
    STD of IQR  & $\sigma \left\{\text{IQR of column}\right\}$ & Data Distribution \\
    \hline
    Mean Q1 & $\mu \left\{\text{Q1 of column}\right\}$ & Data Distribution \\
    \hline
    STD of Q1  & $\sigma \left\{\text{Q1 of column}\right\}$ & Data Distribution \\
    \hline
    Mean Q3  & $\mu \left\{\text{Q3 of column}\right\}$ & Data Distribution \\
    \hline
    STD of Q3  & $\sigma \left\{\text{Q3 of column}\right\}$ & Data Distribution \\
    \hline
    Mean Percent of Z-Score Outliers  & 
    $\mu \left\{ \frac{\text{\# of Z-Score Outliers}}{|\mathbf{A}|_r} \right\}$ & Outlier Data Points \\
    \hline
    STD of Percent of Z-Score Outliers  &
    $\sigma \left\{ \frac{\text{\# of Z-Score Outliers}}{|\mathbf{A}|_r} \right\}$ & Outlier Data Points \\
    \hline
    Mean Correlation& $\mu \left\{\text{Correlation Matrix}\right\}$ & Data linearity \\
    \hline
    STD of Correlation & $\sigma \left\{\text{Correlation Matrix}\right\}$ & Data Linearity \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\subsection{Results}
\begin{table}
  \caption{Experiment Results}
  \label{model-options-table}
  \centering
  \begin{tabular}{llllll}
    \toprule
    % group 2, 3 as multicolumn and 4, 5 as multicolumn
    Dataset & \multicolumn{2}{c}{Accuracy} & \multicolumn{2}{c}{Time to Train (sec)} & Size (RxC)\\
    \cmidrule(lr){2-3} \cmidrule(lr){4-5}
    & SMEML & Dumb & SMEML & Dumb \\
    \midrule
    Lung Cancer Prediction & .9839 & .9839 & 143.07 & 317.04 & 309x16\\
    Chronic Kidney Disease Dataset & 1.0 & 1.0 & 419.64 & 788.46 & 400x26\\
    Pima Indians Diabetes Dataset & .7857 & .8052 & 161.99 & 374.01 & 768x8 \\
    Brain Stroke Prediction Dataset & .9458 & .9458 & 114.16 & 370.95 & 4892x11\\
    Stroke Prediction Dataset & .9393 & .9393 & 172.67 & 385.23 & 5110x12 \\
    Eye State Classification & .9536 & .9549 & 312.57 & 696.51 & 14980x15\\ 
    \bottomrule
  \end{tabular}
\end{table}
\section{Discussion}

>>>>

\begin{table}
  \caption{Survey of Existing AutoML Frameworks}
  \label{automl-survey}
  \centering
  \begin{tabular}{lllllllll}
    \toprule
    Framework & EOU & Algs & FE & TM & HPT & Inter & Custom\\
    \midrule
    H2O AutoML & Easy & 6 & Basic & Yes & Auto & SHAP & High\\
    TPOT & Hard & 10 & Basic & No & Genetic & POJO & Moderate\\
    MLJAR & Easy & 12 & Advanced & Yes & Auto & Basic & High\\
    FLAML & Easy & 9 & Limited & Yes & Auto & SHAP & Low-Moderate\\
    LightAutoML & Easy & 11 & Advanced & Yes & Auto & Basic & Moderate\\
    \bottomrule
  \end{tabular}
  \captionsetup{font=footnotesize}
  \caption*{\textbf{Abbreviations}: EOU: Ease of Use, Algs: Algorithm Types, FE: Feature Engineering, TM: Time Management, HPT: Hyper-Parameter Tuning, Inter: Interface, Custom: Customization}
\end{table}
Table 1: Comparison of six AutoML frameworks based on commonmly evaluated criteria as well as more obscure features relevant to the medical diagnosis process

@@

To implement this model elimination process, we trained a boosting model on 268 kaggle datasets. We chose datasets that were medically related and focused our approach on binary classification. We extracted the attributes in Table 2 from each dataset: this is the input for the boosting model. We then ran a "dumb" framework that trained every model on each dataset. The accuracy of every model is ranked and every model is treated as a label with its value being its rank. Then we trained a multilabel regression to predict the rank of each model, giving the extracted attributes. We produced a median spearman correlation of 0.7301 between the predicted rank and the actual rank and a median p-value of 1.032e-05. Furthermore, we guarantee that we can predict a top two model within the predicted top ten 90.12\% of the time.

When a user inputs a dataset into SMEML, it first extracts the attributes from the dataset. It feeds these attributes into the boosting model to get the predicted rank of each model. Since we have determined that we can predict a top two model within the predicted top ten quite confidently, we can eliminate the bottom 18 models from the list. We run Bayseian optimization on the top ten models to get the best hyperparamters for each model. We select the top models that have significantly better performance than other models. These models are stacked together to form the final model. This process is shown in Figure 1.
\begin{table}
  \caption{Attributes extracted from a dataset}
  \label{model-options-table}
  \centering
  \begin{tabular}{lll}
    \toprule
    Attribute & Formula & Purpose\\
    \midrule
    \# of Rows & $|\mathbf{A}|_r$ & Data Size \\
    \hline
    \# of Columns & $|\mathbf{A}|_c$ & Data Size \\
    \hline
    Target Distribution &
    $\frac{\max(\text{count}_1, \text{count}_0)}{|\mathbf{A}|_r}$
    & Data Type \\
    \hline
    Numerical Columns &
    $\frac{|\mathbf{A}_{num}|}{|\mathbf{A}|_{c}}$
    & Data Type \\
    \hline
    Binary Categorical Columns &
    $\frac{|\mathbf{A}_{bin}|}{|\mathbf{A}|_{c}}$
    & Data Type \\
    \hline
    Mean \# of Distinct Values  & $\mu \left\{\text{\# of distinct values in column}\right\}$ & Data Type \\ 
    \hline
    Mean IQR  & $\mu \left\{\text{IQR of column}\right\}$ & Data Distribution \\ 
    \hline
    STD of IQR  & $\sigma \left\{\text{IQR of column}\right\}$ & Data Distribution \\
    \hline
    Mean Q1 & $\mu \left\{\text{Q1 of column}\right\}$ & Data Distribution \\
    \hline
    STD of Q1  & $\sigma \left\{\text{Q1 of column}\right\}$ & Data Distribution \\
    \hline
    Mean Q3  & $\mu \left\{\text{Q3 of column}\right\}$ & Data Distribution \\
    \hline
    STD of Q3  & $\sigma \left\{\text{Q3 of column}\right\}$ & Data Distribution \\
    \hline
    Mean Percent of Z-Score Outliers  & 
    $\mu \left\{ \frac{\text{\# of Z-Score Outliers}}{|\mathbf{A}|_r} \right\}$ & Outlier Data Points \\
    \hline
    STD of Percent of Z-Score Outliers  &
    $\sigma \left\{ \frac{\text{\# of Z-Score Outliers}}{|\mathbf{A}|_r} \right\}$ & Outlier Data Points \\
    \hline
    Mean Correlation& $\mu \left\{\text{Correlation Matrix}\right\}$ & Data linearity \\
    \hline
    STD of Correlation & $\sigma \left\{\text{Correlation Matrix}\right\}$ & Data Linearity \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\subsection{Results}
\begin{table}
  \caption{Experiment Results}
  \label{model-options-table}
  \centering
  \begin{tabular}{llllll}
    \toprule
    % group 2, 3 as multicolumn and 4, 5 as multicolumn
    Dataset & \multicolumn{2}{c}{Accuracy} & \multicolumn{2}{c}{Time to Train (sec)} & Size (RxC)\\
    \cmidrule(lr){2-3} \cmidrule(lr){4-5}
    & SMEML & Dumb & SMEML & Dumb \\
    \midrule
    Lung Cancer Prediction & .9839 & .9839 & 143.07 & 317.04 & 309x16\\
    Chronic Kidney Disease Dataset & 1.0 & 1.0 & 419.64 & 788.46 & 400x26\\
    Pima Indians Diabetes Dataset & .7857 & .8052 & 161.99 & 374.01 & 768x8 \\
    Brain Stroke Prediction Dataset & .9458 & .9458 & 114.16 & 370.95 & 4892x11\\
    Stroke Prediction Dataset & .9393 & .9393 & 172.67 & 385.23 & 5110x12 \\
    Eye State Classification & .9536 & .9549 & 312.57 & 696.51 & 14980x15\\ 
    \bottomrule
  \end{tabular}
\end{table}

After running the trials, several observations were readily apparent. First, accurcy did not change for 4 of the 6 trials. The other two trials on eye states and diabetes prediction represented a loss of accuracy for SMEML by 0.13\% and 1.95\%, respectively. Training time, however, favored SMEML in every trial. SMEML outperformed the dummy framework by 173.96 seconds in its worst performance on the lung cancer trial and by 383.94 seconds in its best performance on the eye state dataset. Relative to the dummy's performance, a percent decrease in time to run for SMEML of at least 46.78\% could be seen, with a decrease of up to 69.22\% observerd as well. SMEML's worst performance by percent decrease was also the performance in which it returned perfect accuracy. 
\section{Discussion}

As evidenced by our experiment's results, SMEML manages to return comparable accuracy results compared to its dumb couterpart while more than splitting the run time in half in the majority of trials. Its performance is not limited by the size of the dataset, and it in fact returned its first, third, and fourth best percentage-based run times on the 'large' dataset trials (those exceeding 1000 rows). This is important in third world medicine where time is of the essence and there is a constant influx of new patient data. Further research should be done in incorporating this innovation into a full-fledged AutoML framework to expedite and improve the process further. A practical demonsration of this software's real-world application can be found on our SmartDx website.
